Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2023-02-13 analysis/paper/paper.qmd
minor edits while waiting for GP

The TCSA varies between assemblages without any discernible temporal pattern in functional use (@fig-tcsa-radio-carbon panel A). Stemmed points were mainly used before the LGM (@fig-tcsa-radio-carbon panel B). While we predicted the greatest variation in TCSA values during the LGM, our results show that variability decreased during the LGM, in comparison to the previous period. This could be explained by the overall decline in stemmed points usage. We found that stemmed points are only located in certain eco-regional zones such as Central Temperature zone without any clear pattern in TCSA function (@fig-gis-tcsa-sites). However, as predicted, TCSA values in inland areas such as CT of vegetation zones show higher variability. Based on the results, we assume that stemmed points performed a wider range of tasks in low-productivity patches. Both temporal and spatial analyses show that stemmed points were made for a particular environment, not so much harsh as LGM. @prates2022changes claimed that fishtail projectile points in South America was used for megafauna hunting and contributed to their extinction by proving strong relationship between the temporal density and spatial distribution of megafauna and the projectile points. Our results support this claim by showing the dense distribution of stemmed point in certain environment.

Our models premise that the choice of weapon-tip types reflects the socio-environmental circumstances. To examine the possible connections between the roles of stemmed points and environmental change or population dynamics during the Late Paleolithic period, we compared the temporal and spatial patterns of TCSA to results of simulated Mean Annual Temperature (MAT) and Summed probability distribution (SPD) from @park2022did. The simulated temperature distribution in south Korea during the Late Paleolithic period (Figure 9: A from @park2022did) seems similar to the modern eco-regional zones that we used in this research (@fig-gis-tcsa-sites panel A & D). Figure 9: C from @park2022did shows the slow decrease of temperature until the end of LGM, which could explain our results of temporal pattern, meaning stemmed points were used during the cooler environment with a temperature between 4 and 6 ℃. Results of SPD (Figure 10 from @park2022did) show several peaks, as indicated by the positive deviations from the three null models (shaded in red), and downs (shaded in blue), but no drastic growth or decline during our research period, indicating no major impact from population dynamics on stemmed points.

Overall, our temporal and spatial patterns of TCSA show a wide TCSA range throughout the Late Paleolithic period and between eco-regional zones. Especially, we confirmed the widest TCSA range in a single site, Suyanggae (@fig-tcsa-per-sites). As a consequence, our results are consistent with our second scenario in which stemmed points are described as multifunctional tools, suggesting that people created stemmed points in response to unexpected or varying circumstances in their habitats, similar to Clovis points (@eren2022plains).

@@

Considering the importance of stemmed points for the technological transition during the Late Paleolithic in Korea, our research examines the likely uses of stemmed points by asking three research questions: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns of stemmed point use? We applied the tip cross-sectional area (TCSA) metric because it has been used as a ballistically relevant standard to discriminate different likely use classes of projectile points and it requires only a few measurements on a stone artifact (i.e. width and thickness) to compute the metric. We calculated TCSA for a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites. Then we examined the TCSA values with other variables including raw materials, weight, radiocarbon dates, and site locations. Based on the evolutionary approach, we premised that the way of using stemmed points is likely to reflect the socio-environmental circumstances that people encountered and managed. We examined the possible impact of LGM on stemmed points and their distribution on eco-regional maps.

According to the different weapon-delivery systems that summarize TCSA of projectile points from archaeological and ethnographic cases, a majority of stemmed points might be used the most likely as javelin tips and stabbing spear tips. In general, though, we noted a wide range and also differing distribution of TCSA in each assemblage. Therefore, we conclude that stemmed points might play diverse roles during the Late Paleolithic. From our findings, we infer that people were exposed to an unfamiliar situation and used a stemmed point as a multifunctional tool to carry out multiple tasks. We are aware that discriminating the likely use of small projectile points could be arbitrary [@erlandson2014darts]. Since TCSA covers the critical elements of projectiles, flight and penetration dynamics (i.e. increased or decrease by shape of tip and cross section), we still consider it a useful metric for discriminating between weapon tip tools [@hughes1998getting; @sitton2020tip; @lombard2021variation]. As part of our future research, we will conduct use-wear analyses of stemmed points and experimental research in an effort to elucidate the detailed function of stemmed points and technological links between hunter-gatherer groups.

>>>>

Our results shows that TCSA values vary between assemblages with few discernible temporal patterns in function (@fig-tcsa-radio-carbon panel A). Stemmed points were used most frequently before the LGM (@fig-tcsa-radio-carbon panel B). While we predicted the greatest variation in TCSA values during the LGM, our results show that variability decreased during the LGM, in comparison to the previous period. This could be explained by the overall decline in stemmed points usage.

We found that stemmed points are most frequently located in a small number of eco-regional zones such as the Central Temperature zone with few clears pattern in TCSA function across the zones (@fig-gis-tcsa-sites). As predicted, TCSA values in inland areas, such as vegetation zone CT, show higher variability. Based on these results, we assume that stemmed points performed a wider range of tasks in low-productivity patches. Both temporal and spatial analyses show that stemmed points were likely more adapted for particular environments, rather than specific climate contexts. @prates2022changes claimed that fishtail projectile points in South America was used for megafauna hunting and contributed to their extinction by proving strong relationship between the temporal density and spatial distribution of megafauna and the projectile points. We similarly found that stemmed points are more densly distributed in certain environments in South Korea, perhaps also relating to the distribution of fauna.

Overall, our temporal and spatial patterns of TCSA show a wide range of TCSA values throughout the Late Paleolithic period and between eco-regional zones. We confirmed the widest TCSA range in a single site, Suyanggae (@fig-tcsa-per-sites). As a consequence, our results are consistent with our second scenario in which stemmed points are described as multifunctional tools, suggesting that people created stemmed points in response to unexpected or varying circumstances in their habitats, similar to how @eren2022plains explain Clovis points.

@@

Considering the importance of stemmed points for the technological transition during the Late Paleolithic in Korea, our research examines the likely uses of stemmed points by asking three research questions: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips? How diverse were their likely uses? What are the temporal and spatial patterns of stemmed point use? We applied the tip cross-sectional area (TCSA) metric because it has been used as a ballistically relevant standard to discriminate different likely use classes of projectile points and it requires only a few measurements on a stone artifact (i.e. width and thickness) to compute the metric. We calculated TCSA for a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites. Then we examined the TCSA values with other variables including raw materials, weight, radiocarbon dates, and site locations. Drawing on evolutionary theory, we premised that the way of using stemmed points is likely to reflect the socio-environmental circumstances that people encountered and managed. We examined the possible impact of LGM on stemmed points and their distribution across eco-regional zones.

According to the different weapon-delivery systems that can be inferred from TCSA values, the majority of stemmed points from the Korean Palaeolithic were probably used as javelin tips and stabbing spear tips. In general, though, we noted a wide range and also differing distributions of TCSA values in each assemblage. Therefore, we conclude that stemmed points had diverse roles during the Late Paleolithic. We conclude that people frequently encountered unfamiliar situations and used stemmed points as a multifunctional tool to carry out multiple tasks. We are aware that discriminating the likely use of small numbers of projectile points could be arbitrary [@erlandson2014darts]. Since TCSA covers the critical elements of projectiles, flight and penetration dynamics (i.e. increased or decrease by shape of tip and cross section), we still consider it a useful metric for discriminating between weapon tip tools [@hughes1998getting; @sitton2020tip; @lombard2021variation]. As part of our future research, we will conduct use-wear analyses of stemmed points and experimental research in an effort to elucidate the detailed function of stemmed points and technological links between hunter-gatherer groups.
